Spatial Coordination Of Stomatal Patterning Between Leaf Surfaces In Amphistomatous \emph{Arabidopsis thaliana} Incurs No Photosynthetic Advantage

### 2023-10-17 ms/ms.Rmd
revised length-area model and added to ms results

  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage{booktabs} # for tables
---

@@

```{r}
source("r/header.R")
fit = readr::read_rds("objects/fit.rds")
single_surface_results = readr::read_rds("objects/single_surface_results.rds")
single_surface_anova = readr::read_rds("objects/single_surface_anova.rds")
density_anova = readr::read_rds("objects/density_anova.rds")
dual_surface_results = readr::read_rds("objects/dual_surface_results.rds")
correlation_anova = readr::read_rds("objects/correlation_anova.rds")
pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
model_var = readr::read_rds("objects/model_var.rds")

@@

@@

\noindent where $d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023]. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test). We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@pollard_multiple_2005].

@@

Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023]. Stomatal zone area was calculated using Voronoi tessellation (e.g. \autoref{fig:tessellation-example}). The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$. Stomatal length was measured in ImageJ [@schneider_nih_2012].

@@

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_intercellular_1994] using the finite element method (FEM) following @earles_excess_2017. Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The 'coordination advantage' of offset stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

@@

Across all light treatments and leaf surfaces, stomatal length and stomatal area are weakly positively correlated, indicating some support for our hypothesis that stomata are selected for an optimal operational aperture (\autoref{fig:length-area}).

>>>>

  It consists of two paragraphs.
keywords:
  - amphistomy
  - Arabidopsis thaliana
  - CO$_2$ diffusion
  - finite element method
  - optimality
  - photosynthesis
  - stomata
fontsize: 12pt
#spacing: halfline # could also be oneline
#classoptions:
#  - endnotes
bibliography: stomata-spacing.bib
#csl: mycitationstyle.csl
link-citations: yes
output: 
  rticles::oup_article:
    latex_engine: xelatex
#urlcolor: orange
#linkcolor: green
#citecolor: red
header-includes:
 - \usepackage[nomarkers,tablesfirst]{endfloat} # For figures and tables at end
 - \usepackage{lineno} # For line numbering
 - \linenumbers # For line numbering
 - \usepackage{hyperref} # For autoref
 - \renewcommand{\figureautorefname}{Fig.}
 - \usepackage{booktabs} # for tables
 - \usepackage[detect-none]{siunitx} # for \numrange{}{}
 - \sisetup{range-phrase = \text{--}}

---

@@

```{r}
source("r/header.R")
fit = readr::read_rds("objects/fit.rds")
single_surface_results = readr::read_rds("objects/single_surface_results.rds")
single_surface_anova = readr::read_rds("objects/single_surface_anova.rds")
density_anova = readr::read_rds("objects/density_anova.rds")
dual_surface_results = readr::read_rds("objects/dual_surface_results.rds")
correlation_anova = readr::read_rds("objects/correlation_anova.rds")
fit_emtrend = read_rds("objects/fit_emtrends.rds")
pm_pars = readr::read_csv("raw-data/2d-pm-parameters.csv", show_col_types = FALSE)
model_var = readr::read_rds("objects/model_var.rds")

@@

# Summarize slopes of length against stomatal zone area
fit_slopes = gsub(" ", "", print(fit_emtrend, export = TRUE)$summary) |>
  as_tibble() %>%
  set_colnames(gsub(" ", "", colnames(.))) |>
  mutate(light = factor(light, levels = c("low", "medium", "high"))) |>
  arrange(light, surface) |>
  pmap_chr(~ {
    glue("{light} light, {surface} surface: {s}",
         light = ..1, surface = ..2, s = pretty_est(..3, ..4, ..5))
  }) |> 
  str_c(collapse = "; ")

@@

\noindent where $d_i$ is the distance between $\text{stomate}_i$ and its nearest neighbor. We calculated $\mathrm{NNI}$ using the *R* package **spatialEco** version `r packageVersion("spatialEco")` [@evans_spatialeco_2023]. The observed stomatal distribution is dispersed relative to a uniform random distribution if the observed $NNI$ is greater than 95% of the synthetic $\mathrm{NNI}$ values (one-tailed test). We adjusted $P$-values to account for multiple comparisons using the Benjamini-Hochberg [@benjamini_controlling_1995] false discovery rate procedure implemented in the *R* package **multtest** version `r packageVersion("multtest")` [@wong_multiple_2005].

@@

Finally, we examined the relationship between stomatal zone area and stomatal length using a Bayesian linear mixed-effects model fit with the *R* package **brms** version `r fit$version$brms` [@burkner_brms_2017; @burkner_advanced_2018] and *Stan* version `r fit$version$cmdstan` [@stan_development_team_stan_2023]. Stomatal zone area was calculated using Voronoi tessellation (e.g. \autoref{fig:tessellation-example}). The stomatal zone area, $S_\text{area}$, is the region of the leaf surface whose distance to stomate, $S$, is less than the distance to any other stomate, $S$. Stomatal length was measured in ImageJ [@schneider_nih_2012]. We modeled fixed effects of surface, light treatment, stomatal length, and their 2- and 3-way interactions on $\sqrt{S_\text{area}}$. We included random intercepts, random effects of surface, random slopes, and random surface-by-slope interactions within both plant and individual to account for nonindependence of stomata within the same plant or individual. We also modeled residual variance as a function of light treatment. We sampled the posterior distribution from `r dim(fit$fit)[2]` chains with `r dim(fit$fit)[1]` iterations each after `r dim(fit$fit)[1]` warmup iterations. We calculated convergence diagnostics ($\hat{R}$) and effective sample sizes following @vehtari_rank-normalization_2021. We estimated the marginal slope and 95% highest posterior density (HPD) intervals between stomatal length and $\sqrt{S_\text{area}}$ using the *emtrends* function in the *R* package **emmeans** version `r packageVersion('emmeans')` [@lenth_emmeans_2023].

@@

We modeled photosynthesis CO$_2$ assimilation rate using a spatially-explicit two-dimensional reaction diffusion model using a porous medium approximation [@parkhurst_diffusion_1994] using the finite element method (FEM) following @earles_excess_2017. Consider a two-dimensional leaf where stomata occur on each surface in a regular sequence with interstomatal distance $U$. The main outcome we assessed is the advantage of offsetting the position of stomata on each surface compared to have stomata on the same $x$ position on each surface. With these assumptions, by symmetry, we only need to model two stomata, one abaxial and one adaxial, from $x = 0$ to $x = U/2$ and from the adaxial surface at $y = 0$ to the abaxial surface at $y = L$, the leaf thickness. We arbitrarily set the adaxial stomate at $x = 0$ and toggled the abaxial stomata position between $x = U/2$ (offset) or $x = 0$ (below adaxial stomate). The 'coordination advantage' of offset stomatal position on each surface is the photosynthetic rate of the leaf with offset stomata compared to that with stomata aligned in the same $x$ position:

@@

All parameters converged ($\hat{R} < 1.01$) and effective sample sizes were exceeded $10^3$. Across all light treatments and leaf surfaces, stomatal length and stomatal area are weakly positively correlated (\autoref{fig:length-area}). The slope was significantly greater than zero for all abaxial surfaces, but not for the adaxial surface in low and medium light treatments. The estimated marginal slopes and 95% HPD intervals for each combination of light and surface is: `r fit_slopes`.

### 2023-10-17 ms/ms.tex
revised length-area model and added to ms results

\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage{booktabs}

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1
(\protect\hyperlink{ref-evans_spatialeco_2023}{J. S. Evans and Murphy
2023}). The observed stomatal distribution is dispersed relative to a
uniform random distribution if the observed \(NNI\) is greater than 95\%
of the synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg
(\protect\hyperlink{ref-benjamini_controlling_1995}{\textbf{benjamini\_controlling\_1995?}})
false discovery rate procedure implemented in the \emph{R} package
\textbf{multtest} version 2.56.0
(\protect\hyperlink{ref-pollard_multiple_2005}{\textbf{pollard\_multiple\_2005?}}).

@@

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian generalized non-linear multilevel model
fit with the \emph{R} package \textbf{brms} version 2.15.0
(\protect\hyperlink{ref-burkner_brms_2017}{Bürkner 2017},
\protect\hyperlink{ref-burkner_advanced_2018}{2018}) and \emph{Stan}
version 2.27.0
(\protect\hyperlink{ref-stan_development_team_stan_2023}{\textbf{stan\_development\_team\_stan\_2023?}}).
Stomatal zone area was calculated using Voronoi tessellation
(e.g.~\autoref{fig:tessellation-example}). The stomatal zone area,
\(S_\text{area}\), is the region of the leaf surface whose distance to
stomate, \(S\), is less than the distance to any other stomate, \(S\).
Stomatal length was measured in ImageJ
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}).

@@

We modeled photosynthesis CO\(_2\) assimilation rate using a
spatially-explicit two-dimensional reaction diffusion model using a
porous medium approximation
(\protect\hyperlink{ref-parkhurst_intercellular_1994}{\textbf{parkhurst\_intercellular\_1994?}})
using the finite element method (FEM) following
(\protect\hyperlink{ref-earles_excess_2017}{\textbf{earles\_excess\_2017?}}).
Consider a two-dimensional leaf where stomata occur on each surface in a
regular sequence with interstomatal distance \(U\). The main outcome we
assessed is the advantage of offsetting the position of stomata on each
surface compared to have stomata on the same \(x\) position on each
surface. With these assumptions, by symmetry, we only need to model two
stomata, one abaxial and one adaxial, from \(x = 0\) to \(x = U/2\) and
from the adaxial surface at \(y = 0\) to the abaxial surface at
\(y = L\), the leaf thickness. We arbitrarily set the adaxial stomate at
\(x = 0\) and toggled the abaxial stomata position between \(x = U/2\)
(offset) or \(x = 0\) (below adaxial stomate). The `coordination
advantage' of offset stomatal position on each surface is the
photosynthetic rate of the leaf with offset stomata compared to that
with stomata aligned in the same \(x\) position:

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Correlation between paired abaxial and adaxial leaf surfaces. Dashed line indicates no correlation. Weak positive correlations are not significantly different from null simulations. No differences in abaxial-adaxial correlation were observed between light levels according to an analysis of variance ($\alpha$ = 0.05).}
\label{fig:dual-surface}
\end{figure}

Across all light treatments and leaf surfaces, stomatal length and
stomatal area were weakly positively correlated, indicating some support
for our hypothesis that stomata are selected for an optimal operational
aperture (\autoref{fig:length-area}).

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area. Lines of best fit computed using a bayesian generalized non-linear multilevel model. Each light level and leaf surface exhibits a unique, weakly positive relationship between stomatal zone area and length.}
\label{fig:length-area}
\end{figure}

@@

@@

@@

@@

@@

>>>>

\usepackage[nomarkers,tablesfirst]{endfloat}
\usepackage{lineno}
\linenumbers
\usepackage{hyperref}
\renewcommand{\figureautorefname}{Fig.}
\usepackage{booktabs}
\usepackage[detect-none]{siunitx}
\sisetup{range-phrase = \text{--}}

@@

\noindent where \(d_i\) is the distance between \(\text{stomate}_i\) and
its nearest neighbor. We calculated \(\mathrm{NNI}\) using the \emph{R}
package \textbf{spatialEco} version 2.0.1
(\protect\hyperlink{ref-evans_spatialeco_2023}{J. S. Evans and Murphy
2023}). The observed stomatal distribution is dispersed relative to a
uniform random distribution if the observed \(NNI\) is greater than 95\%
of the synthetic \(\mathrm{NNI}\) values (one-tailed test). We adjusted
\(P\)-values to account for multiple comparisons using the
Benjamini-Hochberg
(\protect\hyperlink{ref-benjamini_controlling_1995}{Benjamini and
Hochberg 1995}) false discovery rate procedure implemented in the
\emph{R} package \textbf{multtest} version 2.56.0
(\protect\hyperlink{ref-wong_multiple_2005}{Pollard, Dudoit, and Van Der
Laan 2005}).

@@

Finally, we examined the relationship between stomatal zone area and
stomatal length using a Bayesian linear mixed-effects model fit with the
\emph{R} package \textbf{brms} version 2.20.4
(\protect\hyperlink{ref-burkner_brms_2017}{Bürkner 2017},
\protect\hyperlink{ref-burkner_advanced_2018}{2018}) and \emph{Stan}
version 2.33.1
(\protect\hyperlink{ref-stan_development_team_stan_2023}{Stan
Development Team 2023}). Stomatal zone area was calculated using Voronoi
tessellation (e.g.~\autoref{fig:tessellation-example}). The stomatal
zone area, \(S_\text{area}\), is the region of the leaf surface whose
distance to stomate, \(S\), is less than the distance to any other
stomate, \(S\). Stomatal length was measured in ImageJ
(\protect\hyperlink{ref-schneider_nih_2012}{Schneider, Rasband, and
Eliceiri 2012}). We modeled fixed effects of surface, light treatment,
stomatal length, and their 2- and 3-way interactions on
\(\sqrt{S_\text{area}}\). We included random intercepts, random effects
of surface, random slopes, and random surface-by-slope interactions
within both plant and individual to account for nonindependence of
stomata within the same plant or individual. We also modeled residual
variance as a function of light treatment. We sampled the posterior
distribution from 4 chains with 1000 iterations each after 1000 warmup
iterations. We calculated convergence diagnostics (\(\hat{R}\)) and
effective sample sizes following Vehtari et al.
(\protect\hyperlink{ref-vehtari_rank-normalization_2021}{2021}). We
estimated the marginal slope and 95\% highest posterior density (HPD)
intervals between stomatal length and \(\sqrt{S_\text{area}}\) using the
\emph{emtrends} function in the \emph{R} package \textbf{emmeans}
version 1.8.8 (\protect\hyperlink{ref-lenth_emmeans_2023}{Lenth 2023}).

@@

We modeled photosynthesis CO\(_2\) assimilation rate using a
spatially-explicit two-dimensional reaction diffusion model using a
porous medium approximation
(\protect\hyperlink{ref-parkhurst_diffusion_1994}{Parkhurst 1994}) using
the finite element method (FEM) following Earles et al.
(\protect\hyperlink{ref-earles_excess_2017}{2017}). Consider a
two-dimensional leaf where stomata occur on each surface in a regular
sequence with interstomatal distance \(U\). The main outcome we assessed
is the advantage of offsetting the position of stomata on each surface
compared to have stomata on the same \(x\) position on each surface.
With these assumptions, by symmetry, we only need to model two stomata,
one abaxial and one adaxial, from \(x = 0\) to \(x = U/2\) and from the
adaxial surface at \(y = 0\) to the abaxial surface at \(y = L\), the
leaf thickness. We arbitrarily set the adaxial stomate at \(x = 0\) and
toggled the abaxial stomata position between \(x = U/2\) (offset) or
\(x = 0\) (below adaxial stomate). The `coordination advantage' of
offset stomatal position on each surface is the photosynthetic rate of
the leaf with offset stomata compared to that with stomata aligned in
the same \(x\) position:

@@

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/dual-surface.pdf}
\caption{Pixel-wise correlation between near stomatal distance (NSD) squared on paired abaxial and adaxial leaf surfaces. Dashed line indicates zero correlation. Weak positive correlations are not significantly different from zero after correcting for multiple comparisons. The correlation does not differ among light treatments.}
\label{fig:dual-surface}
\end{figure}

\hypertarget{larger-stomata-supply-larger-mesophyll-volumes}{%
\subsection{Larger stomata supply larger mesophyll
volumes}\label{larger-stomata-supply-larger-mesophyll-volumes}}

All parameters converged (\(\hat{R} < 1.01\)) and effective sample sizes
were exceeded \(10^3\). Across all light treatments and leaf surfaces,
stomatal length and stomatal area are weakly positively correlated
(\autoref{fig:length-area}). The slope was significantly greater than
zero for all abaxial surfaces, but not for the adaxial surface in low
and medium light treatments. The estimated marginal slopes and 95\% HPD
intervals for each combination of light and surface is: low light,
abaxial surface: 1.928 {[}\numrange{0.779}{3.133}{]}; low light, adaxial
surface: 1.745 {[}\numrange{-0.041}{3.373}{]}; medium light, abaxial
surface: 1.085 {[}\numrange{0.328}{1.957}{]}; medium light, adaxial
surface: 0.656 {[}\numrange{-0.399}{1.691}{]}; high light, abaxial
surface: 0.597 {[}\numrange{0.316}{0.911}{]}; high light, adaxial
surface: 1.269 {[}\numrange{0.831}{1.721}{]}.

\begin{figure}[ht]
\includegraphics[width = \textwidth]{figures/length-area.pdf}
\caption{Stomatal length and stomatal zone area are positively correlated. Linear regression lines and 95\% confidence ribbons are from a Bayesian linear mixed-effects model.}
\label{fig:length-area}
\end{figure}

@@

\leavevmode\vadjust pre{\hypertarget{ref-benjamini_controlling_1995}{}}%
Benjamini, Yoav, and Yosef Hochberg. 1995. {``Controlling the False
Discovery Rate: A Practical and Powerful Approach to Multiple
Testing.''} \emph{Journal of the Royal Statistical Society. Series B
(Methodological)} 57 (1): 289--300.

@@

\leavevmode\vadjust pre{\hypertarget{ref-earles_excess_2017}{}}%
Earles, J. Mason, Guillaume Théroux-Rancourt, Matthew E. Gilbert, Andrew
J. McElrone, and Craig R. Brodersen. 2017. {``Excess Diffuse Light
Absorption in Upper Mesophyll Limits {CO}\(_{\textrm{2}}\) Drawdown and
Depresses Photosynthesis.''} \emph{Plant Physiology} 174 (2): 1082--96.
\url{https://doi.org/10.1104/pp.17.00223}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-lenth_emmeans_2023}{}}%
Lenth, Russell V. 2023. \emph{Emmeans: {Estimated} {Marginal} {Means},
Aka {Least}-{Squares} {Means}}.
\url{https://CRAN.R-project.org/package=emmeans}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-wong_multiple_2005}{}}%
Pollard, K. S., S. Dudoit, and M. J. Van Der Laan. 2005. {``Multiple
{Testing} {Procedures}: The Multtest {Package} and {Applications} to
{Genomics}.''} In \emph{Bioinformatics and {Computational} {Biology}
{Solutions} {Using} {R} and {Bioconductor}}, edited by Wing Wong, M.
Gail, K. Krickeberg, A. Tsiatis, J. Samet, Robert Gentleman, Vincent J.
Carey, Wolfgang Huber, Rafael A. Irizarry, and Sandrine Dudoit, 249--71.
New York, NY: Springer New York.
\url{https://doi.org/10.1007/0-387-29362-0_15}.

@@

\leavevmode\vadjust pre{\hypertarget{ref-stan_development_team_stan_2023}{}}%
Stan Development Team. 2023. \emph{Stan {Modeling} {Language} {Users}
{Guide} and {Reference} {Manual}}. \url{https://mc-stan.org}.

\leavevmode\vadjust pre{\hypertarget{ref-vehtari_rank-normalization_2021}{}}%
Vehtari, Aki, Andrew Gelman, Daniel Simpson, Bob Carpenter, and
Paul-Christian Bürkner. 2021. {``Rank-{Normalization}, {Folding}, and
{Localization}: {An} {Improved} {Rˆ} for {Assessing} {Convergence} of
{MCMC} (with {Discussion}).''} \emph{Bayesian Analysis} 16 (2).
\url{https://doi.org/10.1214/20-BA1221}.
